Latent Variable Clustering for Classification of Events in the AT-TPC

### 2020-06-12 doc/Manuscript/article.tex
Updates from Overleaf

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. %Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of neural networks to clustering by implementing a mixture of autoencoders algorithm, with improved results on real data. %
%However, the neural network clustering suffers from stability issues. %necessitating further investigations into this approach.

%In this work we introduce the application of convolutional autoencoder neural networks to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on ${}^{46}$Ar. 
%The data we analyze were recorded by an active target time-projection chamber (AT-TPC). Machine learning presents an interesting avenue for researchers operating an AT-TPC, as traditional analysis methods of AT-TPC data are both computationally expensive and fit all particle tracks against the event type of interest. The latter presents a considerable challenge when the space of reactions is not known prior to the analysis.

%We explore the performance of the autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well. On simulated data these classifiers achieve an $f1$ score \cite{Chinchor1992} of $f1>0.95$. The VGG16 latent classifier achieves this result with as few as $N=100$ samples, as does the convolutional autoencoder when trained on the VGG16 representations of the particle tracks. On real data, pre-processed with noise filtering, the same models achieve an $f1>0.7$. For unfiltered real data the models achieve an $f1$ value larger than $0.6$. Both of the previous results were found with the classifiers trained on $N=100$ samples. Furthermore, we found that the autoencoder model reduces the variability in the identification of proton events by $64\%$ from the benchmark logistic regression classifier trained on the VGG16 latent space on real experimental data.

@@

The MIXAE algorithm clusters data by pairing a set of encoder-decoder neural networks (autoencoder) with an auxiliary assignment network: each autoencoder constructs a representation of a given event while the assignment network assigns a set of cluster probabilities to the event. {\color[orange]{clear sentences about what the model does}} As noted by \citet{Zhang}, the MIXAE model assumes that the clusters are roughly similar in volume in the data and that one has some reasonable estimate on the number of clusters present {\color{cyan} [Ryan: Is this true of our data? I think if these assumptions are going to be mentioned, then we need to say that our data meets the criteria.}. The architecture is portrayed in figure \ref{fig:mixae}, tapered boxes denote a direction of compression in the network components.

@@

\noindent For a more formal consideration of these heuristics see \citet{Zhang}.

@@

\section{Results and Discussions}\label{sec:results}
\begin{itemize}
\item Clear out the "methods" like parts.
\item merge 5.1 and 5.2 - keeping point 1 in mind.
\item merge 5.3 5.4 and 5.6 and create table from figure 10, 11 and 12. 
\item go over 5.7 - mention instability in MIXAE results, peak performance of MIXAE is good - consistency is not. 
\item remove fig. 14 and fig 15.
\end{itemize}

>>>>

\begin{abstract}
% Modified by DB (6/8/2020)
This article presents novel applications of machine learning methods to the problem of event separation in an active target detector, the Active-Target Time Projection Chamber (AT-TPC) \cite{Bradt2017}. 
The overarching goal is to identify classes of events in the early stages of data processing, before 
%Machine learning presents an interesting avenue for researchers operating the AT-TPC, as traditional analysis methods of AT-TPC data are computationally expensive and have to fit all tracks against the event type of interest. The latter presents a considerable challenge when either the space of reactions is not known prior to the analysis, or the event type of interest is not the most abundant. 
The application of unsupervised clustering  algorithms to the analysis of two-dimensional projections of particle tracks from a resonant proton scattering experiment on $^{46}$Ar is introduced. We explore the performance of autoencoder neural networks and a pre-trained VGG16 \cite{Simonyan2014} convolutional neural network.
%on two tasks: a semi-supervised classification task and the unsupervised clustering of particle tracks. 
%On the semi-supervised task, we find that a logistic regression classifier trained on small labelled subsets of the latent space of these models perform very well, both on simulated and real data. On the clustering task, 
We find that a $K$-means algorithm applied to the simulated data in the VGG16 latent space forms almost perfect clusters. Additionally, the VGG16+K-means approach finds high purity clusters of proton events for real experimental data. We also explore the application of neural networks to clustering by implementing a mixture of autoencoders algorithm, with improved results on real data.

@@

The MIXAE algorithm clusters data by pairing a set of encoder-decoder neural networks (autoencoder) with an auxiliary assignment network: each autoencoder constructs a representation of a given event while the assignment network assigns a set of cluster probabilities to the event. {\color{orange}[clear sentences about what the model does]} . The architecture is portrayed in figure \ref{fig:mixae}, tapered boxes denote a direction of compression in the network components.

@@

\noindent For a more formal consideration of these heuristics and the assumptions made on the data by this model see \citet{Zhang}.

@@

\section{Results and Discussions}\label{sec:results}
